### 2010-03-29 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: fix spacing of operators used in text.

We propose to embed inductive types as a universe in our dependent
type theory, the universe of \emph{descriptions}
(Figure~\ref{fig:desc_universe}). As expected, the code of this
universe mimics the standard grammar of data-types definitions in
simply-typed languages. Hence, we have a $\DSigma{}{}$ code,
interpreted as a $\Sigma$-type, to build the
sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
contains no useful payload. The functoriality of the data-types is
introduced by $\DIndx$. When we tie the knot with a fix-point, the
hole left open by $\DIndx$ will be turned into a standard recursive
argument. We notice that this functoriality appears in the type of
$\descop{\V{D}}{}$ itself, for a given data-type definition $\V{D}$. This
corresponds to the object part of the functor described by $\V{D}$.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DUnit          & \Desc \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
    \DIndx          & \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}ll@{\:=\:\:}ll}
\descop{\DUnit}{& \V{X}}        &  \Unit                                       \\
\descop{\DSigma{\V{S}}{\V{D}}}{& \V{X}} &  \SIGMAS{\V{s} : \V{S}}{\descop{\V{D}~\V{s}}{\V{X}}}         \\
\descop{\DIndx{\V{D}}}{& \V{X}}     &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
\end{array}
}
\end{array}
}\]

@@

\[\stk{
\NatD : \Desc \\
\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{} ]})}
                     {[ \DUnit \quad (\DIndx{\DUnit}) ]}
}\]

Let us explain its construction. First, we use $\DSigma{}{}$ to give a
choice between the $\NatZero$ and $\NatSuc{}$ constructors. In the
$\NatZero$ case, we reach the end of the description: there is no
useful payload in that case. In the $\NatSuc{}$ case, we left one hole
open for the recursive argument, and we close the description.

@@

\[\stk{
\ListD : \Set \To \Desc \\
\ListD \: \V{X} \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons ]})}
                           {[ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) ]}
}\]

The modification consists in turning the $\NatSuc{}$ constructor into
a proper $\ListCons$ taking an argument in $\V{X}$ followed by a recursive
argument. In this case, we use $\DSigma{}{}$ in its product
interpretation: we pack an element of $\V{X}$ together with the recursive
argument. Easily, one sees that this code actually describes the
following functor:

@@

From the examples above, we observe that data-types are defined by a
$\DSigma$ which first argument is a finite set of constructors. The
descriptions fitting into this pattern are called \emph{tagged}
description. Formally, we have:

@@

So far, for convenience, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its code, $\DSigma{}{}$, $\DIndx{}$,
and $\DUnit$ as type formers, extending the basic type theory. Hence,
in an implementation of this type theory, we need to extended it with
these constructors and their typing rules. Although it makes no
conceptual difference, the code of description should rather be read
as a specification. We promise the existence of such objects,
satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\All{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                 \PITEL{\V{X}}{\Set}
                 \PITEL{\V{P}}{\V{X} \To \Set} \\
               & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                 \Set 
\end{array} \\
\All{\DId}{\V{X}}{\V{P}}{\V{x}} \mapsto \V{P}\: \V{x} \\
\All{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{z}} \mapsto \Unit \\
\All{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\pair{\V{d}}{\V{d'}}{}} \mapsto \TIMES{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}{\All{\V{D'}}{\V{X}}{\V{P}}{\V{d'}}} \\
\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} \mapsto \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{b}} \\
\All{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{f}} \mapsto \PI{\V{\V{s}}}{\V{S}} \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{(\V{f}\: \V{s})}
}
&
%%
\stk{
\begin{array}{@{}ll}
\all{}{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X}}{\Set}
                   \PITEL{\V{P}}{\V{X} \To \Set} \\
                 & \PITEL{\V{R}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
\end{array} \\
\all{\DId}{\V{X}}{\V{P}}{\V{R}}{\V{x}} = \V{R}\: \V{x} \\
\all{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{R}}{\V{z}} = \Void \\
\all{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{d}}{\V{d'}}{}} = \pair{\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{d}}}{\all{\V{D'}}{\V{X}}{\V{P}}{\V{R}}{\V{d'}}}{} \\
\all{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} = \all{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
\all{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\V{f}} = \LAM{\V{s}} \all{(\V{T}\: \V{s})}{\V{X}}{\V{P}}{\V{R}}{(\V{f}\: \V{s})}
\end{array}
}
\]

>>>>

We propose to embed inductive types as a universe in our dependent
type theory, the universe of \emph{descriptions}
(Figure~\ref{fig:desc_universe}). As expected, the code of this
universe mimics the standard grammar of data-types definitions in
simply-typed languages. Hence, we have a $\DSigma{\!}{\!}$ code,
interpreted as a $\Sigma$-type, to build the
sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
contains no useful payload. The functoriality of the data-types is
introduced by $\DIndx$. When we tie the knot with a fix-point, the
hole left open by $\DIndx$ will be turned into a standard recursive
argument. We notice that this functoriality appears in the type of
$\descop{D}{}$ itself, for a given data-type definition $D$. This
corresponds to the object part of the functor described by $D$.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DUnit          & \Desc \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DIndx          & \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}ll@{\:=\:\:}ll}
\descop{\DUnit}{& \V{X}}        &  \Unit                                       \\
\descop{\DSigma{\V{S}}{\V{D}}}{& \V{X}} &  \SIGMAS{\V{s} : \V{S}}{\descop{\V{D}~\V{s}}{\V{X}}}         \\
\descop{\DIndx{\V{D}}}{& \V{X}}     &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
\end{array}
}
\end{array}
}\]

@@

\[\stk{
\NatD : \Desc \\
\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
                     {[ \DUnit \quad (\DIndx{\DUnit}) ]}
}\]

Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
give a choice between the $\NatZero$ and $\NatSuc{\!}$ constructors. In
the $\NatZero$ case, we reach the end of the description: there is no
useful payload in that case. In the $\NatSuc{\!}$ case, we left one hole
open for the recursive argument, and we close the description.

@@

\[\stk{
\ListD : \Set \To \Desc \\
\ListD \: X \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons ]})}
                           {[ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{\DUnit}}) ]}
}\]

The modification consists in turning the $\NatSuc{}$ constructor into
a proper $\ListCons$ taking an argument in X followed by a recursive
argument. In this case, we use $\DSigma{}{}$ in its product
interpretation: we pack an element of $\V{X}$ together with the recursive
argument. Easily, one sees that this code actually describes the
following functor:

@@

From the examples above, we observe that data-types are defined by a
$\DSigma{\!}{\!}$ which first argument is a finite set of
constructors. The descriptions fitting into this pattern are called
\emph{tagged} description. Formally, we have:

@@

So far, for convenience, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its code, $\DSigma{\!}{\!}$,
$\DIndx{\!}$, and $\DUnit$ as type formers, extending the basic type
theory. Hence, in an implementation of this type theory, we need to
extended it with these constructors and their typing rules. Although
it makes no conceptual difference, the code of description should
rather be read as a specification. We promise the existence of such
objects, satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\All{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                 \PITEL{\V{X}}{\Set}
                 \PITEL{\V{P}}{\V{X} \To \Set} \\
               & \PI{\V{xs}}{\descop{D}{\V{X}}} 
                 \Set 
\end{array} \\
\All{\DId}{\V{X}}{\V{P}}{\V{x}} \mapsto \V{P}\: \V{x} \\
\All{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{z}} \mapsto \Unit \\
\All{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\pair{\V{d}}{\V{d'}}{}} \mapsto \TIMES{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}{\All{\V{D'}}{\V{X}}{\V{P}}{\V{d'}}} \\
\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} \mapsto \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{b}} \\
\All{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{f}} \mapsto \PI{\V{\V{s}}}{\V{S}} \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{(\V{f}\: \V{s})}
}
&
%%
\stk{
\begin{array}{@{}ll}
\all{}{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X}}{\Set}
                   \PITEL{\V{P}}{\V{X} \To \Set} \\
                 & \PITEL{\V{R}}{\PI{\V{x}}{X} P\: x}
                   \PI{\V{xs}}{\descop{D}{\V{X}}} 
                   \All{D}{X}{P}{xs} 
\end{array} \\
\all{\DId}{\V{X}}{\V{P}}{\V{R}}{\V{x}} = \V{R}\: \V{x} \\
\all{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{R}}{\V{z}} = \Void \\
\all{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{d}}{\V{d'}}{}} = \pair{\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{d}}}{\all{\V{D'}}{\V{X}}{\V{P}}{\V{R}}{\V{d'}}}{} \\
\all{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} = \all{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
\all{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\V{f}} = \LAM{\V{s}} \all{(\V{T}\: \V{s})}{\V{X}}{\V{P}}{\V{R}}{(\V{f}\: \V{s})}
\end{array}
}
\]

### 2010-03-29 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix spacing of operators used in text.

The names $\NilE$ and $\ConsE{}{}$ are absolutely not a
coincidence. $\EnumU$ is indeed isomorphic to a list of $\UId$s. In
Section~\ref{sec:desc-examples}, we have shown how to build lists in
the universe of description. Therefore, the actual implementation of
$\EnumU$ boils down to the following, trivial definition:

@@

Another interesting consequence concerns the $\spi{}{}$ operator: it
can actually be implemented with the generic $\F{induction}$
principle. From an implementation perspective, this means that we
remove an operator hard-coded in the host language. This operator is
simply implemented in the dependently-typed target language, as any
other function.

@@

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{} ]})  \\
                           & \left[\begin{array}{l}
                                   \DUnit                                \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \SHED)}      \\
                                   \DIndx{\DUnit}                        \\
                                   \end{array}
                             \right]
\end{array}
}\]

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's constructors:
$\DUnit$, $\DSigma$, and $\DIndx$. The reader will notice that the
tagged notation we have used so far for $\Desc$ constructors now fully
make sense: these were actually the tags we are defining. \note{This
  sentence badly needs a rephrasing.}

The case of $\DUnit$ is trivial: we immediately reach the end of the
data-type. $\DUnit$ has no payload. The case of $\DIndx{}$ comes
naturally: it only takes a recursive argument -- a description. The
case of $\DSigma$ is problematic. Recall the specification of
$\DSigma$:

\[    \DSigma{}{} : \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc      \]

@@

In order to capture a notion of higher-order induction, we need to
extend our universe of descriptions. This consists in adding a code
$\DHindx$ that takes an indexing set $\V{H}$. The code and its
interpretation are presented in
Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx$ use the elements
of $H$ to index as many recursive arguments.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \ldots          & \:\:\ldots \\
    \DHindx         & \PI{\V{H}}{\Set} \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\ldots                        &  \ldots \\
\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     &  \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}}
\end{array}
}
\end{array}
}\]

@@

Note that $\DIndx$ is isomorphic to $\DHindx{\Unit}$. However, we
tolerate this duplication and do not replace $\DIndx$ by the more
general $\DHindx$. Indeed, unlike its counterpart, $\DIndx$ is
first-order. Therefore, while both codes are \emph{extensionally} the
same, in most practical implementation they would be dealt with rather
differently. While definitional equality can cope with first order
objects, the functional presentation introduced by $\DHindx$ is
unlikely to be amenable to a purely definitional treatment.

Equipped with $\DHindx$, we can describe the extended code of our
universe of data-types, as shown in Figure~\ref{fig:desc-levitate}.
The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
successfully describe the $\DSigma$ case, by a simple appeal to the
higher-order induction on $\V{S}$. The $\DHindx$ case consists in packing
the $\V{H}$ in $\Set$ with a recursive argument.

@@

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{}, \DHindx{}{} ]}) \\
                           & \left[\begin{array}{l}
                                   \DUnit                                            \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{\V{S}}{\DUnit})}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set}{(\LAM{\V{H}} \DIndx{\DUnit})}
                                   \end{array}
                             \right]
\end{array}
}\]

@@

\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
    \pair{\ListCons{\DVar{}}{\V{E}}}
         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
}\]

We simply add a constructor, $\DVar{}$, and define its argument to be
a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep $\V{E}$ and $\V{D}$
as they were, hence leaving the functor unchanged. Unfolding the
interpretation of this definition, we convince ourself that this
corresponds to the functor $\FMFreeMonadD$. The fix-point operation
ties the knot and gives us the full-blown free monad construction.

@@

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})}) \To
           \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\etag{t}\,}{\V{ys}}{} & \Con{\pair{\etag{t}}{\V{ys}}{}}
\end{array}
}\]

>>>>

The names $\NilE$ and $\ConsE{\!}{\!}$ are absolutely not a
coincidence. $\EnumU$ is indeed isomorphic to a list of $\UId$s. In
Section~\ref{sec:desc-examples}, we have shown how to build lists in
the universe of description. Therefore, the actual implementation of
$\EnumU$ boils down to the following, trivial definition:

@@

Another interesting consequence concerns the $\spi{\!}{\!}$ operator:
it can actually be implemented with the generic $\F{induction}$
principle. From an implementation perspective, this means that we
remove an operator hard-coded in the host language. This operator is
simply implemented in the dependently-typed target language, as any
other function.

@@

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!} ]})  \\
                               & \left[\begin{array}{l}
                                   \DUnit                                \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \SHED)}      \\
                                   \DIndx{\DUnit}                        \\
                                 \end{array}
                                 \right]
\end{array}
}\]

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's constructors:
$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice that the
tagged notation we have used so far for $\Desc$ constructors now fully
make sense: these were actually the tags we are defining. \note{This
  sentence badly needs a rephrasing.}

The case of $\DUnit$ is trivial: we immediately reach the end of the
data-type. $\DUnit$ has no payload. The case of $\DIndx{\!}$ comes
naturally: it only takes a recursive argument -- a description. The
case of $\DSigma{\!}{\!}$ is problematic. Recall the specification of
$\DSigma{\!}{\!}$:

\[    \DSigma{}{} : \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc      \]

@@

In order to capture a notion of higher-order induction, we need to
extend our universe of descriptions. This consists in adding a code
$\DHindx$ that takes an indexing set $H$. The code and its
interpretation are presented in
Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx{\!}{\!}$ use the
elements of $H$ to index as many recursive arguments.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \ldots          & \:\:\ldots \\
    \DHindx         & \PI{H}{\Set} \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\ldots                        &  \ldots \\
\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     &  \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}}
\end{array}
}
\end{array}
}\]

@@

Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
we tolerate this duplication and do not replace $\DIndx{\!}$ by the
more general $\DHindx{\!}{\!}$. Indeed, unlike its counterpart,
$\DIndx{\!}$ is first-order. Therefore, while both codes are
\emph{extensionally} the same, in most practical implementation they
would be dealt with rather differently. While definitional equality
can cope with first order objects, the functional presentation
introduced by $\DHindx{\!}{\!}$ is unlikely to be amenable to a purely
definitional treatment.

Equipped with $\DHindx$, we can describe the extended code of our
universe of data-types, as shown in Figure~\ref{fig:desc-levitate}.
The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
successfully describe the $\DSigma$ case, by a simple appeal to the
higher-order induction on $S$. The $\DHindx$ case consists in packing
the $H$ in $\Set$ with a recursive argument.

@@

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!}, \DHindx{\!}{\!} ]}) \\
                           & \left[\begin{array}{l}
                                   \DUnit                                            \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{\V{S}}{\DUnit})}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set}{(\LAM{\V{H}} \DIndx{\DUnit})}
                                   \end{array}
                             \right]
\end{array}
}\]

@@

\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{E}{D}{}}\:X \mapsto
    \pair{\ListCons{\DVar{}}{E}}
         {\pair{\DSigma{X}{\DUnit}}{D}{}}{}
}\]

We simply add a constructor, $\DVar{}$, and define its argument to be
a $\DSigma{X}{\DUnit}$, that is an element of $X$. We keep $E$ and $D$
as they were, hence leaving the functor unchanged. Unfolding the
interpretation of this definition, we convince ourself that this
corresponds to the functor $\FMFreeMonadD$. The fix-point operation
ties the knot and gives us the full-blown free monad construction.

@@

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})}) \To
           \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}   & \sigma\: x                   \\
\apply\: D\: X\: Y\: \sigma\: \pair{\etag{t}}{ys}{} & \Con{\pair{\etag{t}}{ys}{}}
\end{array}
}\]

### 2010-03-29 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: Tidied up the code in the paper, fixed a few transcription bugs and added colour (well, purple)
Desc@ICFP: fix spacing of operators used in text.

As presented, our type theory suffers from a major weakness. Indeed,
we are subject to Girard's paradox, as we assume that $\Set$ lives in
$\Set$. We made that choice for presentational convenience, as
universe stratification is orthogonal to our work. Nonetheless, our
universes of description rather naturally leads itself to
stratification. Unsurprisingly, $\IDesc{}$ at level $l$ is of type
$\Set^{\blue{l+1}}$. Similarly, the interpretation of $\IDesc{}$ at
level $l$ is an object of type $\Set^{\blue{l}}$:

\[\stk{
\data \IDesc{\!}^{\blue{l}} (\Bhab{\V{I}}{\Set^{\blue{l+1}}}) : \Set^{\blue{l+1}} \where \\
\;\;\ldots \\
\\
\idescop{\_\:}{}{}^{\blue{l}} : \PI{\V{I}}{\Set^{\blue{l+1}}} \IDesc{{\!}^{\blue{l}}I} \To (\V{I} \To \Set^{\blue{l}}) \To \Set^{\blue{l}}    \\
\ldots
}\]

Consequently, we can implement the operations and examples developed
above. We refer the reader to our Agda implementation, which take
advantage of set polymorphism to implement the universe of indexed
descriptions at any level. Further, we have coded $\IDesc{}$ in itself
and have proved the isomorphism between the host and the embedded
universes.

@@

However, we differ in several ways. First, they adopt a generative
perspective: each universe extends the base type theory with both type
formers and elimination rules. Thanks to the levitation, we only rely
on a generic induction and a specialised $\switchD{}{}{}$. Second, the
authors develop several universes, each with its own strengths and
weaknesses, but lack a formal, unifying universe. With indexed
descriptions, we subsume simple descriptions, hence avoiding the move
to and from several codings. Finally, the authors often resort to an
extensional notion of equality, while we have given an
equality-agnostic presentation. Moreover, our presentation is arranged
so as to use definitional equality as much as possible. Hence, in
practice, the developer is relieved from many trivial proof
obligations.

>>>>

As presented, our type theory suffers from a major weakness. Indeed,
we are subject to Girard's paradox, as we assume that $\Set$ lives in
$\Set$. We made that choice for presentational convenience, as
universe stratification is orthogonal to our work. Nonetheless, our
universes of description rather naturally leads itself to
stratification. Unsurprisingly, $\IDesc{\!}$ at level $l$ is of type
$\Set^{\blue{l+1}}$. Similarly, the interpretation of $\IDesc{\!}$ at
level $l$ is an object of type $\Set^{\blue{l}}$:

\[\stk{
\data \IDesc{\!}^{\blue{l}} (\Bhab{\V{I}}{\Set^{\blue{l+1}}}) : \Set^{\blue{l+1}} \where \\
\;\;\ldots \\
\\
\idescop{\_\:}{}{}^{\blue{l}} : \PI{\V{I}}{\Set^{\blue{l+1}}} \IDesc{{\!}^{\blue{l}}\V{I}} \To (\V{I} \To \Set^{\blue{l}}) \To \Set^{\blue{l}}    \\
\ldots
}\]

Consequently, we can implement the operations and examples developed
above. We refer the reader to our Agda implementation, which take
advantage of set polymorphism to implement the universe of indexed
descriptions at any level. Further, we have coded $\IDesc{\!}$ in
itself and have proved the isomorphism between the host and the
embedded universes.

@@

However, we differ in several ways. First, they adopt a generative
perspective: each universe extends the base type theory with both type
formers and elimination rules. Thanks to the levitation, we only rely
on a generic induction and a specialised
$\switchD{\!}{\!}{\!}$. Second, the authors develop several universes,
each with its own strengths and weaknesses, but lack a formal,
unifying universe. With indexed descriptions, we subsume simple
descriptions, hence avoiding the move to and from several
codings. Finally, the authors often resort to an extensional notion of
equality, while we have given an equality-agnostic
presentation. Moreover, our presentation is arranged so as to use
definitional equality as much as possible. Hence, in practice, the
developer is relieved from many trivial proof obligations.
